Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-08-10 Paper/Manuscript.tex
Vivek paper added to Refs and mentioned in footnote 2.
Some minor corrections

However, it is not clear how much this applies to BNS systems.}.

@@

Given Moore's law and new search approaches (e.g., see Ref.~\cite{Gabbard:2017lja} for one based on deep learning) 
we speculate that the advances in the detection, localization and early-warning algorithms in the next decade will reduce the required advance
warning time to less than an hour; thus we set $T_\text{AW} = 1\,\text{hour}$ and obtain corresponding BNS horizon distances of $D_H=87$ and $613\,$Mpc ($z\approx 0.02,0.127$) for ET-B and ET-C, respectively.
In other words, all inspiralling BNSs closer than $D_H$ will give us more than an hour's leeway before the GRB (employing the detection criterion $\text{SNR} =15$ as before).
The volume set by $D=D_H$ translates to a BNS merger rate of $\sim\mathcal{O}(1)$ for ET-B and
$\approx 355^{+730}_{-280}\,\text{year}^{-1}$ for ET-C. Additionally, each source within $D_H$ will accumulate $\bar\rho_F \gtrsim 420, 58$ over its inspiral
in the ET-B, ET-C bands, respectively. These results are summarized in Table~\ref{table:horizon}.

Both Tables~\ref{table:ET} and \ref{table:horizon} clearly show that ET will yield superb SNRs for sources out to 1\,Gpc. 
With such high SNRs we expect that a future network consisting of ET and LIGO Voyager will be able to localize half the sources 
at $D\approx 1000\,$Mpc to within $ 10\,\text{deg}^2$ and more than half the sources at $D\lesssim 400\,$Mpc to $\approx 1\,\text{deg}^2$ \cite{Mills:2017urp}.
Therefore, we optimistically conclude that 2030s hold in store real-time observations of dozens to hundreds of gamma-ray bursts
\footnote{As the ET construction/commissioning timeline is currently uncertain, we can only speculate that ET-C will be operational after 2030,
which is roughly in line with the milli-Hertz space interferometer LISA's timeline \cite{Audley:2017drz}.}.
%It is also reasonable to speculate that by that time the detection, localization and early-warning algorithms will have become sufficiently sophisticated
%to alert the electromagnetic observatories within an hour of the initial detection (SNR = 15). 
%
We summarize our main findings for both the ALV network and ET-B/C in Fig.~\ref{fig:summary}.
%
%
%
\begin{table}[h]
\centering
\begin{tabular}{l|ccc}
%toprule
\hline
 & ET-B & & ET-C\T\B\\
\hline
  $D_H $& 87\,Mpc& &{613\,Mpc}\T\\
  $R(D_H) $& $1^{+2}_{-1}\,\text{yr}^{-1}$&\hspace{5mm} &{$355^{+730}_{-280}\,\text{yr}^{-1}$}\T\\
  $\bar\rho_F(D_H)$ & 420 &&{58}\T\B\\
\hline
\end{tabular}
\caption{Horizon distances of ET-B and ET-C assuming $T_\text{AW} =1\,$hour. $R(D_H)$ is the BNS merger rate within a volume of $D_H^3$
obtained by rescaling the rate inferred from Advanced LIGO's O1, O2 observing periods \cite{GW170817}. $\bar\rho_F(D_H)$ is the total SNR accumulated due to a BNS inspiralling at $D_H$ [see Eq.~(\ref{eq:rhoF_ET})].}\label{table:horizon}
\end{table}
%
%
%
\begin{figure}[ht!]
\includegraphics[width=\linewidth]{../Figures/Main_results.pdf}
\caption{Summary of our major findings for both the Advanced-LIGO-Virgo (ALV) network circa 2020 and the Einstein Telescope's B and C configurations
ca. 2030. The light coloured (gray) data shows $T_\text{AW}$ for the ALV2020 (dotted), ET-B (dashed), and ET-C (solid), respectively.
The connecting light (gray) lines are drawn only for visualization purposes unlike the dark (blue) lines coming from the $D^{-1}$ powerlaw of the SNR [see Eq.~(\ref{eq:ET_SNRv2})] which is plotted on the right axis. 
We used $D=50,100,200\,$Mpc to obtain the ALV results (Sec.~\ref{Sec:ALIGO2020}) and $D=100,200,400,1000\,$Mpc for the ET results.}\label{fig:summary}
\end{figure}
%
%
%
% LISA does not see these sources, not sensitive enough!
%
%
\section{Binary black hole neutron star systems}\label{sec:BH_NS}
In this section, we investigate the possibility of witnessing the tidal distruption of a neutron star by a black hole before the merger.
The treatment of the inspiral is the same as before, however the fate of the neutron star depends strongly on the system's intrinsic parameters
such as the black hole mass and spin, and the neutron star's equation of state (EOS) and compactness $c_\text{NS}\equiv G M_\text{NS}/(R_\text{NS}c^2)$. 
The evolution of black hole-neutron star (BH-NS) binaries is a very active field of research at the interface of numerical relativity and high-energy astrophysics
(see Ref. \cite{Shibata:2011jka} for a comprehensive review) requiring fully general relativistic magnetohydrodynamic treatment including neutrino transport equations.
The complexity of evolving the dynamics of these systems require large computational resources and long runtimes.
%running on large-scale clusters.
As such, the details are beyond the scope of this article, but we can use a mostly Newtonian treatment to serve our purposes. %, however the numerical simulationshave verified

A necessary, but insufficient, condition for tidal disruption (TD) is that
\be
\f{2 G M_\text{BH} (c_R R_\text{NS})}{r^3} \gtrsim \f{G M_\text{NS}}{(c_R R_\text{NS})^2} \label{eq:TD_cond1},
\ee
where $r$ is the orbital radius, $R_\text{NS}$ is the radius of the NS and $M$ denotes masses. 
$c_R R_\text{NS}$ is the semi-major axis of the elongated oblate spheroid that represents the tidally distorted shaped of the NS to leading order with
$c_R >1 $. %Setting $m_2= M_\text{BH}, m_1 = M_\text{NS}$, and $r_1=r_\text{NS}$, 
The above conditions roughly implies
%
\be
\f{M_\text{BH}}{M_\text{NS}} \gtrsim \left(\f{r}{R_\text{NS}}\right)^3 \label{eq:TD_cond2}.
\ee
%
We wish to maximize the chances of observing a TD event, therefore we ingrain the BH-NS binary with certain desirable features
some of which can be deduced from Eqs.~(\ref{eq:TD_cond1}, \ref{eq:TD_cond2}). These are
%
%
\begin{enumerate}[(i)]
 \item Small separation. As the Newtonian tidal force due to the black hole is proportional to $M_\text{BH}/r^3$, $r$ has to be
 minimized to provide maximum tidal force. 
 However, for $r< r_\text{ISCO}$ the NS simply plunges into the BH; therefore the minium separation is given by the ISCO radius.
 \item Low black hole mass. In the test-mass case, $r_\text{ISCO} \propto M_\text{BH}$, so the Newtonian tidal force roughly scales as $M_\text{BH}/r_\text{ISCO}^3\propto M_\text{BH}^{-2}$. 
 Hence, $M_\text{BH}$ must also be minimized. 
 As is the case with $r$, there is also a lower bound to $M_\text{BH}$ set from astrophysical observations and stellar evolution models, which currently yield $M_\text{BH} \gtrsim 5 M_\odot$ \cite{Farr:2010tu, Raithel:2017nlc, Ozel:2010su, Wiktorowicz:2013dua, Christian:2018mjv}.
 \item High black hole spins. $r_\text{ISCO}$ is much smaller in the case of a prograde NS orbit around a spinning Kerr black hole.
 For instance, for the maximum dimensionless BH spin value of $a\equiv c J_\text{BH}/(G M^2_\text{BH})=1$, $r_\text{ISCO}$ for prograde orbits is 1/6th of the Schwarzschild value $6GM_\text{BH}/c^2$ whereas for $a=1$ retrograde orbits, $r_\text{ISCO}$ is 3/2 of the Schwarzschild value.
 With smaller ISCO separation, there is much more tidal force exerted on the NS before it plunges in.
 \item Less compact neutron stars. Eq.~(\ref{eq:TD_cond1}) can be rearranged to isolate $c_\text{NS}$
 on the right hand side. Smaller values for $c_\text{NS}$ mean that TD is more likely to occur for $r>r_\text{ISCO}$.
\end{enumerate}
%

@@

There is additional motivation to choose low BH mass as 
 $\tau_\text{insp} \propto M_c^{-5/3} = (1+Q)^{1/3}/Q$ for $Q\equiv M_\text{BH}/M_\text{NS} > 1$, which scales as $Q^{-2/3}$ as $Q\to\infty$ , i.e., heavier binaries inspiral faster
 so it is desirable to minimize $Q$.
 Setting $M_\text{NS}=1.4 M_\odot$ as before gives $Q=5/1.4  \simeq 3.571$. $c_\text{NS}$ is determined by the radius of the NS constrained to be in the $9.9 - 11.2\,$km range \footnote{Recent work by the ALV collaboration \cite{GW170817_radii}) give bounds of $R_\text{NS}=9.1 -11.8\,$km, more or less consistent with astrophysical results.}\cite{Ozel:2016oaf}
 which results in $0.13\lesssim c_\text{NS} \lesssim 0.15$ 
 for our chosen NS mass. 
%As we wish to avoid discussing the details of the strong-field evolution of the binary, we borrow certain results from the literature to
%justify our choice for the chosen parameters above. 
If we pick the more pessimistic $c_\text{NS} = 0.15$ bound then we must have $Q\lesssim 5$ and $a\gtrsim 0.5$ (prograde) for TD to occur \cite{Ferrari:2008nr}
and $Q< 3$ in the case of $a=0$ \cite{Kyutoku:2011vz}.

>>>>

However, it is not clear how much this applies to BNS systems, nonetheless see Ref.~\cite{Chaurasia:2018zhg} for a recent numerical study of an eccentric BNS merger.}.

@@

Given Moore's law and new search approaches (e.g., see Ref.~\cite{Gabbard:2017lja} for one based on deep learning) 
we speculate that the advances in the detection, localization and early-warning algorithms in the next decade will reduce the required advance
warning time to less than an hour; thus we set $T_\text{AW} = 1\,\text{hour}$ and obtain corresponding BNS horizon distances of $D_H=87$ and $613\,$Mpc ($z\approx 0.02,0.127$) for ET-B and ET-C, respectively.
In other words, all inspiralling BNSs closer than $D_H$ will give us more than an hour's leeway before the GRB (employing the detection criterion $\text{SNR} =15$ as before).
The volume set by $D=D_H$ translates to a BNS merger rate of $\sim\mathcal{O}(1)\,\text{yr}^{-1}$ for ET-B and
$\approx 355^{+730}_{-280}\,\text{yr}^{-1}$ for ET-C. Additionally, each source within $D_H$ will accumulate $\bar\rho_F \gtrsim 420, 58$ over its inspiral
in the ET-B, ET-C bands, respectively. These results are summarized in Table~\ref{table:horizon}.

Both Tables~\ref{table:ET} and \ref{table:horizon} clearly show that ET will yield superb SNRs for sources out to 1\,Gpc. 
With such high SNRs we expect that a future network consisting of ET and LIGO Voyager will be able to localize half the sources 
at $D\approx 1000\,$Mpc to within $ 10\,\text{deg}^2$ and more than half the sources at $D\lesssim 400\,$Mpc to $\approx 1\,\text{deg}^2$ \cite{Mills:2017urp}.
Therefore, we optimistically conclude that 2030s hold in store real-time observations of dozens to hundreds of gamma-ray bursts
\footnote{As the ET construction/commissioning timeline is currently uncertain, we can only speculate that ET-C will be operational after 2030,
which is roughly in line with the milli-Hertz space interferometer LISA's timeline \cite{Audley:2017drz}.}.
%It is also reasonable to speculate that by that time the detection, localization and early-warning algorithms will have become sufficiently sophisticated
%to alert the electromagnetic observatories within an hour of the initial detection (SNR = 15). 
%
We summarize our main findings for both the ALV network and ET-B/C in Fig.~\ref{fig:summary}.
%
%
%
\begin{table}[h]
\centering
\begin{tabular}{l|ccc}
%toprule
\hline
 & ET-B & & ET-C\T\B\\
\hline
  $D_H $& 87\,Mpc& &{613\,Mpc}\T\\
  $R(D_H) $& $1^{+2}_{-1}\,\text{yr}^{-1}$&\hspace{5mm} &{$355^{+730}_{-280}\,\text{yr}^{-1}$}\T\\
  $\bar\rho_F(D_H)$ & 420 &&{58}\T\B\\
\hline
\end{tabular}
\caption{Horizon distances of ET-B and ET-C assuming $T_\text{AW} =1\,$hour. $R(D_H)$ is the BNS merger rate within a volume of $D_H^3$
obtained by rescaling the rate inferred from Advanced LIGO's O1, O2 observing periods \cite{GW170817}. $\bar\rho_F(D_H)$ is the total SNR accumulated due to a BNS inspiralling at $D_H$ [see Eq.~(\ref{eq:rhoF_ET})].}\label{table:horizon}
\end{table}
%
%
%
\begin{figure}[ht!]
\includegraphics[width=\linewidth]{../Figures/Main_results.pdf}
\caption{Summary of our major findings for both the Advanced-LIGO-Virgo (ALV) network circa 2020 and the Einstein Telescope's B and C configurations
ca. 2030. The light coloured (gray) data shows $T_\text{AW}$ for the ALV2020 (dotted), ET-B (dashed), and ET-C (solid), respectively.
The connecting light (gray) lines are drawn only for visualization purposes unlike the dark (blue) lines coming from the $D^{-1}$ powerlaw of the SNR [see Eq.~(\ref{eq:ET_SNRv2})] which is plotted on the right axis. 
We used $D=50,100,200\,$Mpc to obtain the ALV results (Sec.~\ref{Sec:ALIGO2020}) and $D=100,200,400,1000\,$Mpc for the ET results.}\label{fig:summary}
\end{figure}
%
%
%
% LISA does not see these sources, not sensitive enough!
%
%
\section{Binary black hole neutron star systems}\label{sec:BH_NS}
In this section, we investigate the possibility of witnessing the tidal distruption of a neutron star by a black hole before the merger.
The treatment of the inspiral is the same as before, however the fate of the neutron star depends strongly on the system's intrinsic parameters
such as the black hole mass and spin, and the neutron star's equation of state (EOS) and compactness $\mathcal{C}_\text{NS}\equiv G M_\text{NS}/(R_\text{NS}c^2)$. 
The evolution of black hole-neutron star (BH-NS) binaries is a very active field of research at the interface of numerical relativity and high-energy astrophysics
(see Ref. \cite{Shibata:2011jka} for a comprehensive review) requiring fully general relativistic magnetohydrodynamic treatment including neutrino transport equations.
The complexity of evolving the dynamics of these systems require large computational resources and long runtimes.
%running on large-scale clusters.
As such, the details are beyond the scope of this article, but we can use a mostly Newtonian treatment to serve our purposes. %, however the numerical simulationshave verified

A necessary, but insufficient, condition for tidal disruption (TD) is that
\be
\f{2 G M_\text{BH} (c_R R_\text{NS})}{r^3} \gtrsim \f{G M_\text{NS}}{(c_R R_\text{NS})^2} \label{eq:TD_cond1},
\ee
where $r$ is the orbital radius, $R_\text{NS}$ is the radius of the NS and $M$ denotes masses. 
$c_R R_\text{NS}$ is the semi-major axis of the elongated oblate spheroid that represents the tidally distorted shaped of the NS to leading order with
$c_R >1 $. %Setting $m_2= M_\text{BH}, m_1 = M_\text{NS}$, and $r_1=r_\text{NS}$, 
The above conditions roughly implies
%
\be
\f{M_\text{BH}}{M_\text{NS}} \gtrsim \left(\f{r}{R_\text{NS}}\right)^3 \label{eq:TD_cond2}.
\ee
%
We wish to maximize the chances of observing a TD event, therefore we ingrain the BH-NS binary with certain desirable features
some of which can be deduced from Eqs.~(\ref{eq:TD_cond1}, \ref{eq:TD_cond2}). These are
%
%
\begin{enumerate}[(i)]
 \item Small separation. As the Newtonian tidal force due to the black hole is proportional to $M_\text{BH}/r^3$, $r$ has to be
 minimized to provide maximum tidal force. 
 However, for $r< r_\text{ISCO}$ the NS simply plunges into the BH; therefore the minium separation is given by the ISCO radius.
 \item Low black hole mass. In the test-mass case, $r_\text{ISCO} \propto M_\text{BH}$, so the Newtonian tidal force roughly scales as $M_\text{BH}/r_\text{ISCO}^3\propto M_\text{BH}^{-2}$. 
 Hence, $M_\text{BH}$ must also be minimized. 
 As is the case with $r$, there is also a lower bound to $M_\text{BH}$ set from astrophysical observations and stellar evolution models, which currently yield $M_\text{BH} \gtrsim 5 M_\odot$ \cite{Farr:2010tu, Raithel:2017nlc, Ozel:2010su, Wiktorowicz:2013dua, Christian:2018mjv}.
 \item High black hole spins. $r_\text{ISCO}$ is much smaller in the case of a prograde NS orbit around a spinning Kerr black hole.
 For instance, for the maximum dimensionless BH spin value of $a\equiv c J_\text{BH}/(G M^2_\text{BH})=1$, $r_\text{ISCO}$ for prograde orbits is 1/6th of the Schwarzschild value $6GM_\text{BH}/c^2$ whereas for $a=1$ retrograde orbits, $r_\text{ISCO}$ is 3/2 of the Schwarzschild value.
 With smaller ISCO separation, there is much more tidal force exerted on the NS before it plunges in.
 \item Less compact neutron stars. Eq.~(\ref{eq:TD_cond1}) can be rearranged to isolate $\mathcal{C}_\text{NS}$
 on the right hand side. Smaller values for $\mathcal{C}_\text{NS}$ mean that TD is more likely to occur for $r>r_\text{ISCO}$.
\end{enumerate}
%

@@

There is additional motivation to choose low BH mass as 
 $\tau_\text{insp} \propto M_c^{-5/3} = (1+Q)^{1/3}/Q$ for $Q\equiv M_\text{BH}/M_\text{NS} > 1$, which scales as $Q^{-2/3}$ as $Q\to\infty$ , i.e., heavier binaries inspiral faster
 so it is desirable to minimize $Q$.
 Setting $M_\text{NS}=1.4 M_\odot$ as before gives $Q=5/1.4  \simeq 3.571$. $\mathcal{C}_\text{NS}$ is determined by the radius of the NS constrained to be in the $9.9 - 11.2\,$km range \footnote{Recent work by the ALV collaboration \cite{GW170817_radii}) give bounds of $R_\text{NS}=9.1 -11.8\,$km, more or less consistent with astrophysical results.}\cite{Ozel:2016oaf}
 which results in $0.13\lesssim \mathcal{C}_\text{NS} \lesssim 0.15$ 
 for our chosen NS mass. 
%As we wish to avoid discussing the details of the strong-field evolution of the binary, we borrow certain results from the literature to
%justify our choice for the chosen parameters above. 
If we pick the more pessimistic $\mathcal{C}_\text{NS} = 0.15$ bound then we must have $Q\lesssim 5$ and $a\gtrsim 0.5$ (prograde) for TD to occur \cite{Ferrari:2008nr}
and $Q< 3$ in the case of $a=0$ \cite{Kyutoku:2011vz}.
